---
title: "XAI Practice - PDP for Bike Rentals and Houses"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r}

library(tidyverse)
library(randomForest)
library(pdp)
library(caret)
library(ggplot2)
library(dplyr)
```

# 1.Bike Rentals Prediction

## 1.1 Load and preprocess data

```{r}
bike_data <- read.csv("C:/Users/Usuario/OneDrive/Documentos/day.csv")

# Apply transformations as indicated in the instructions
bike_data <- bike_data %>%
  mutate(
    season_spring = ifelse(season == 1, 1, 0),
    season_summer = ifelse(season == 2, 1, 0),
    season_fall = ifelse(season == 3, 1, 0),
    MISTY = ifelse(weathersit == 2, 1, 0),
    RAIN = ifelse(weathersit >= 3, 1, 0),
    temp_denorm = temp * 47 - 8,
    hum_denorm = hum * 100,
    windspeed_denorm = windspeed * 67,
    days_since_2011 = as.integer(as.Date(dteday) - as.Date("2011-01-01"))
  )
```

## 1.2 Fit Random Forest model

```{r}
set.seed(123)
# Define predictor variables including dummy variables
vars_rf <- c("season_spring", "season_summer", "season_fall",
             "MISTY", "RAIN", "temp_denorm", "hum_denorm", "windspeed_denorm",
             "days_since_2011")

rf_model <- randomForest(cnt ~ ., data = bike_data[, c("cnt", vars_rf)], ntree = 200)
print(rf_model)
```

## 1.3 One-dimensional Partial Dependence Plots (PDP)

```{r}
features_pdp <- c("days_since_2011", "temp_denorm", "hum_denorm", "windspeed_denorm")

for (feat in features_pdp) {
  p <- partial(rf_model, pred.var = feat, grid.resolution = 30)
  print(
    autoplot(p) + 
      ggtitle(paste("PDP for", feat)) +
      xlab(feat) + ylab("Predicted bike rentals")
  )
}
```

## 1.4 Variable Importance Plot

```{r}
varImpPlot(rf_model)
```

## 1.5 ICE Plot for days_since_2011

```{r}
library(pdp)
ice_days <- partial(rf_model, pred.var = "days_since_2011", ice = TRUE)
autoplot(ice_days)

```

## 1.6 Two-dimensional Partial Dependence Plot (Temperature and Humidity)

```{r}
# Sample randomly to reduce computational cost
set.seed(123)
bike_sample <- bike_data %>% sample_n(500)

pdp_2d <- partial(rf_model, pred.var = c("temp_denorm", "hum_denorm"), 
                  grid.resolution = 30, progress = "text", chull = TRUE, 
                  train = bike_sample)

# 2D PDP using geom_tile
ggplot(pdp_2d, aes(x = temp_denorm, y = hum_denorm, fill = yhat)) +
  geom_tile(width = 1, height = 1) + 
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "2D PDP: Temperature and Humidity", 
       x = "Temperature (denormalized)", y = "Humidity (%)", fill = "Predicted rentals") +
  theme_minimal()
```

## 1.7 Joint density plot of Temperature and Humidity in the dataset

```{r}
ggplot(bike_data, aes(x = temp_denorm, y = hum_denorm)) +
  geom_point(alpha = 0.3, color = "grey") +
  geom_density_2d(color = "blue") +
  labs(title = "Joint Density of Temperature and Humidity")

```

# 2. House Price Prediction

## 2.1 Load data and sample

```{r}
house_data <- read.csv("C:/Users/Usuario/OneDrive/Documentos/kc_house_data.csv")

# Random sample to speed up computations
set.seed(123)
house_sample <- house_data %>% sample_n(10000)

# Variables to use as predictors
vars_house <- c("bedrooms", "bathrooms", "sqft_living", "sqft_lot", "floors", "yr_built")
```

## 2.2 Fit Random Forest model for house prices

```{r}
rf_house <- randomForest(price ~ ., data = house_sample[, c("price", vars_house)], ntree = 200)
print(rf_house)
```

## 2.3 One-dimensional Partial Dependence Plots for house features

```{r}
library(pdp)
library(ggplot2)

features_house_pdp <- c("bedrooms", "bathrooms", "sqft_living", "floors")

pdp_house_list <- lapply(features_house_pdp, function(feat) {
  p <- partial(rf_house, pred.var = feat, grid.resolution = 20)
  autoplot(p) + 
    ggtitle(paste("PDP for", feat)) +
    xlab(feat) + ylab("Predicted price")
})

for (plot_obj in pdp_house_list) {
  print(plot_obj)
}
```

## 2.4 Variable Importance Plot for house price model

```{r}
varImpPlot(rf_house)
```

## 2.5 ICE Plot for sqft_living

```{r}
ice_living <- partial(rf_house, pred.var = "sqft_living", ice = TRUE)
autoplot(ice_living)
```
